Attributions for Poverty in Post-Socialist Countries

### 2014-02-14 article2013.Rmd
Kirjoitin teoriaa ja tutkimusasetelmaa ajantasaiseksi. Jatkan lounaan jälkeen tuloksista

---
title: Attributions for Poverty in Post-Socialist Countries
author: [Markus Kainu & Mikko Niemelä]
date: February 12, 2014
abstract: Utilising The Life in Transition Survey (LiTS) conducted in 2010, the purpose of the paper is to examine public perceptions of the causes of poverty in Europe and Western Asia with special emphasis on post-socialist countries. The study applies multilevel techniques in order to analyse the role of individual and country-level explanatory factors as determinants of poverty attributions. As country-level determinants, paper analyses the changes in country's economic performance during the period of global financial crises, income inequality and the state of democracy. As individual-level determinants, the paper tests to what extent risk factors related to financial hardship are associated with perceptions of the causes of poverty.
keywords: poverty
output:
  pdf_document:
    toc: true
    number_section: true
---

@@

Earlier studies have shown that attributions for poverty differ in general between the West and the East of Europe but the variation between East Central European countries is relatively large. There is, however, a shortage of comparative studies which have focused particularly on differences within post-socialist countries. In addition, most of the prior research is based on attitude surveys from the 1990s and most of them have restricted country selection. Therefore, as previous studies have suggested (Kallio and Niemelä, 2013; Lepianka et al., 2010), there is a need for detailed and more up-to-date analysis which would focus on the attributions for poverty within the post-socialist space. Hence, the general purpose of the study is to analyze public perceptions of the causes of poverty in Europe and Western Asia with special emphasis on post-socialist countries.

@@

As country-level determinants of perceptions, we will analyze, whether this variation can be explained by different regimes within post-socialist countries and the post-socialist legacy. The study divides post-socialist countries into two groups: Central Eastern Europe (CEE) and Commonwealth of Independent States (CIS). CEE includes Czech Republic, Estonia, Hungary, Bulgaria, Latvia, Lithuania, Poland, Slovakia, Slovenia, Romania; and CIS[^****] Armenia, Azerbaijan, Belarus, Georgia, Kazakhstan, Kyrgyzstan, Moldova, Tajikistan, Ukraine, Uzbekistan and Russia. The survey data enables us also to compare post-socialist countries to a selection of Western European nations (France, Germany, Italy and Sweden). Thus, in order to examine the distinctiveness of post-socialist perception, Western European countries are included in the descriptive analyses. (See Figure 1.)

@@

[^****]:Georgia withdrawn its membership from CIS in 2008

Besides the country groups we will also analyze the role of changes in country’s economic performance during the period of global financial crises, income inequality and the state of democracy. As described in previous sections, we assume that in those countries where the economic crises has meant declining economic conditions, public are more likely to endorse social explanations for poverty than in countries where national economy has continued to grow despite the global economic turbulence. Based on the theoretical discussion on civil liberties, political rights and institutional trust, we assume that the public support to social blame type of explanation is high in those countries where the state of democracy is low. Due to contradictory findings on the direction of the effect of income inequality, it is more difficult to provide any straightforward hypothesis on the importance of income inequality.

@@

Effect of financial crises        Whether respondents perceives that their household has suffered the financial crisis great or fair amount     list

@@

An additional dataset of selected macro indicators were added to the LiTS2-dataset consisting of four variables: 1) change in total GDP in 2007 - 2010, 2) gini coefficient of economic inequality, 3) 2011 democracy index and 4) country grouping. Data for change in GDP is from World Bank's [World Development Indicators](http://data.worldbank.org/data-catalog/world-development-indicators), gini coefficients from [UNU-WIDER World Income Inequality Database](http://www.wider.unu.edu/research/Database/en_GB/database/) and Democracy Index from the [Economist Intelligence Unit](http://www.eiu.com/Default.aspx). Gini coefficient for each country is the latest available varying from 2002 to 2006. (see Appendix A)

`Voice and accountability captures perceptions of the extent to which a country's citizens are able to participate in selecting their government, as well as freedom of expression, freedom of association, and a free media. [info.worldbank.org/governance/wgi/pdf/va.pdf](http://info.worldbank.org/governance/wgi/pdf/va.pdf) - *leikkasin tulos-luvusta tän kappaleen*`

@@

The empirical section starts by presenting the results of descriptive analyses. As we assume that individual level perceptions are dependent both on individual and country-level contextual factors, it is recommended to use multilevel regression analysis for this task (Snijders \& Bosker, 2011).  Multilevel analysis requires hierarchical dataset and allows the estimation of country-level effects on individual level outcomes. The analysis is performed using lmer-function from lme4 package in R. All of the models employed are so called multilevel logistic random intercept models at the level of individuals and countries.

Our multilevel modelling focuses first in bivariate connections between two dependent variables and both individual and country level independent variables(See Appendix B). At the second step we include all individual level variables to a single model (models sb1 and ib1). Then we add each of the four country level variables one at the time (models sb2/ib2 - sb5/ib5). With limited number of countries we are obliged to use only few contextual variables. Tables will include both a fixed effects and a random effects part. The fixed effects part will include odds ratios and statistical significances of independent variables, while the random effects part will indicate the country-level variance and standard deviation, deviance and number of cases and groups.

>>>>

---
title: Attributions for Poverty in Post-Socialist Countries
author: [Markus Kainu & Mikko Niemelä]
date: February 14, 2014
abstract: Utilising The Life in Transition Survey (LiTS) conducted in 2010, the purpose of the paper is to examine public perceptions of the causes of poverty in Europe and Western Asia with special emphasis on post-socialist countries. The study applies multilevel techniques in order to analyse the role of individual and country-level explanatory factors as determinants of poverty attributions. As country-level determinants, paper analyses the changes in country's economic performance during the period of global financial crises, income inequality and the state of democracy. As individual-level determinants, the paper tests to what extent risk factors related to financial hardship are associated with perceptions of the causes of poverty.
keywords: poverty
output:
  pdf_document:
    toc: true
    number_section: true
---

@@

Earlier studies have shown that attributions for poverty differ in general between the West and the East of Europe but the variation between East Central European countries is relatively large. There is, however, a shortage of comparative studies which have focused particularly on differences within post-socialist countries. In addition, most of the prior research is based on attitude surveys from the 1990s and most of them have restricted country selection. Therefore, as previous studies have suggested (Kallio and Niemelä, 2013; Lepianka et al., 2010), there is a need for detailed and more up-to-date analysis which would focus on the attributions for poverty within the post-socialist space. Hence, the general purpose of the study is to analyze public perceptions of the causes of poverty in Europe and Central Asia with special emphasis on post-socialist countries.

@@

We target our analyses at 22 post-socialist countries and use five Western European countries as references in descriptive analyses (see figure 1.). Our selection on 22 post-socialist countries includes Czech Republic, Estonia, Hungary, Bulgaria, Latvia, Lithuania, Poland, Slovakia, Slovenia, Romania from Central Eastern Europe (CEE), and Armenia, Azerbaijan, Belarus, Georgia, Kazakhstan, Kyrgyzstan, Moldova, Tajikistan, Ukraine, Uzbekistan and Russia from Commonwealth of Independent States (CIS)[^***]. Of Western European countries we have included France, Germany, Italy, United Kingdom and Sweden as references.

@@

[^***]:Georgia withdrawn its membership from CIS in 2008

As contextual level predictors we use country’s economic performance during the period of global financial crises, living conditions and the state of democracy. As described in previous sections, we assume that in those countries where the economic crises has meant declining economic conditions, public are more likely to endorse social explanations for poverty than in countries where national economy has continued to grow despite the global economic turbulence. Based on the theoretical discussion on civil liberties, political rights and institutional trust, we assume that the public support to social blame type of explanation is high in those countries where the state of democracy is low. The case with human deveplopment in less clear, but we assume, against the hypothesis by Lepianka et al. (2010), that lower level of human development and lack of welfare institutions may well be connected with higher support for individual blame type of explanation.

@@

Effect of financial crises        Whether respondents perceives that their household has suffered the financial crisis great or fair amount     lits

@@

An additional dataset of selected macro indicators were added to the LiTS2-dataset consisting of three variables: 1) change in total GDP in 2007 - 2010, 2) Human Deveplopment Index in 2010 and 3) Voice and accountability variable. Data for change in GDP is from World Bank's [World Development Indicators](http://data.worldbank.org/data-catalog/world-development-indicators), Human Development Index from [United NAtions Development Programme](http://hdr.undp.org/en/data) and Voice and Accountability from the World Bank's [Worldwide Governance Indicators](http://info.worldbank.org/governance/wgi/index.aspx#home). All macro level indicators are obtained from [Quality of Government Institutes](http://www.qog.pol.gu.se/) data using [rQog](http://markuskainu.fi/rqog/)-package (Kainu 2014).

<!-- seuraava kappale piilossa

@@

-->

The empirical section starts by presenting the results of descriptive analyses. As we assume that individual level perceptions are dependent both on individual and country-level contextual factors and we have a hierarchical two level dataset, it is recommended to use multilevel regression analysis for this task (Snijders & Bosker, 2011) as multilevel regression analysis allows the estimation of country-level effects on individual level outcomes. The analysis is performed using glmer-function from lme4 package in R. All of the models employed are so called multilevel logistic random intercept models at the level of individuals and countries.

We are modelling the date in stepwise manner first running the empty models for both social blame (table 3) and individual blame (table 4) type of explanation. At second stage we include all individual level predictors and in third stage we include only the macro level predictors. In final models we include both the individual and macro level variables. We assume that our dependent variables are independent, but to confirm this we run a further model where we predict support for social blame against the support for individual blame using both the individual and macro level variables. This model is compared with full models from tables 3 and 4 in table 5.

`Tables will include both a fixed effects and a random effects part. The fixed effects part will include odds ratios and statistical significances of independent variables, while the random effects part will indicate the country-level variance and standard deviation, deviance and number of cases and groups.`
